Peptide retention time prediction

### 2015-01-26 retention.tex
Small textual changes

\begin{wrapfigure}{r}{.45\textwidth}
\vspace{-10pt}
%\begin{figure}[htbp]
\centering
\includegraphics[trim=0.5cm 0cm 2cm 1.5cm, clip=true, width=0.43\textwidth]{img/reproducibility.pdf}
\caption{\label{fig:repr} {\bf Reproducibility of peptide retention time.} Two identical shotgun proteomics experiments were carried out. The retention times of the peptides confidently identified in both runs are displayed.}
%\end{figure}
\vspace{-25pt}
\end{wrapfigure}

% iRT? \cite{Escher:2012qv}
Following these observations, a variety of retention time prediction
methods has been proposed. Most naturally one could divide these
methods into four classes: look-up tables-based approaches,
index-based methods, modeling-based methods and machine learning
techniques. The main ideas underlying each category are described
below, while Table \ref{tab:rtmethods} lists a few of the most popular
retention time prediction methods from each category.

@@

Moruz {\em et al.} \cite{elude1} developed a software package for
retention time prediction which we named {\sc Elude}. One of the main
novelty points of {\sc Elude} was the calculation of peptide features
based on both a general hydrophobicity index, as well as a retention
index trained for each dataset at hand. This ensured that the
algorithm could be applied to different chromatography setups with
minimal loss in performance. For cases when little training data was
available, {\sc Elude} implemented an additional workflow based on a
library of pre-trained retention models. This workflow required only a
few tens of training peptides to select and calibrate the most
suitable model from the library. Both the selection and calibration
were performed using statistical methods robust to outliers.

However, despite all these efforts, it should be pointed out that none
of the retention time prediction methods currently available is able
to provide accuracy comparable to the small shifts observed when
running a sample twice through a chromatography column
(Figure \ref{fig:repr}). Recently, the lack of suitable
representations for secondary structure elements was indicated as one
of the main causes of prediction errors \cite{Reimer2012}.  Most
probably additional factors are involved, and further endeavors are
required to reach the desired level of accuracy.

@@

To address this, in Moruz {\em et al.}\cite{elude2} extended the first
version of their algorithm {\sc ELUDE} to encompass modified peptides
as well. This was straightforward to achieve once the authors observed
that the in-house trained retention index contained nearly all the
information incorporated in the more general hydrophobicity scale.
This meant that they could remove all the features based on this index
with minimal loss in accuracy. Further, they redefined all features
describing a peptide in terms of the in-house retention index alone.
Since this index was trained for each dataset at hand, the algorithm
became fully trainable for any modified amino acid (Figure
\ref{fig:elude}). The new version of {\sc ELUDE} was evaluated on
peptides containing artifactual modifications introduced as part of an
MS protocol called COmbined FRActional DIagonal Chromatography
(COFRADIC) \cite{Gevaert2002}. The results showed that {\sc ELUDE}
yielded equally accurate predictions for modified and unmodified
peptides \cite{elude2}.

@@

% how it works and main results
More specifically, we matched observed masses and retention times of
unfragmented ions to predicted masses and retention times for the
theoretical peptides from an {\it in silico} digest. We built a
statistical framework to assign protein probabilities based on these
matches, and combined the results with the ones obtained by the
typical workflow based on fragmented ions. Following this approach, we
obtained up to 7\% more protein identifications at a fixed false
discovery rate threshold of 1\% \cite{mf}.

The idea of identifying proteins based on entire peptides, rather than
peptide fragments, is not new. In fact, this was the principle
underlying peptide mass fingerprinting, one of the first methods used
to identify proteins via mass spectrometry. However, when introduced
in the 90's, mass fingerprinting could not be applied to unequivocally
identify proteins in complex biological samples. Consequently, shotgun
proteomics has gained popularity instead. Nevertheless, with great
developments in instrumentation leading to sub-ppm mass accuracy,
the possibility of using mass fingerprinting-like methods came back to
attention~\cite{iamt1}. Our study showed how improved mass accuracy
and retention time predictions enable the use of unfragmented ions
even in the context of complex peptide mixtures.

@@

% Improve the design of SRM experiments 
In addition, predicted retention times are useful investigating
the elution properties of the peptides.  To illustrate this, we predicted the retention time of all the
theoretical peptides from an {\it in silico} digest of the human
proteome using  {\sc ELUDE}  \cite{elude1}. This exercise revealed a rather skewed
distribution, with a majority of the peptides eluting in only a small
portion of the chromatography run. This suggested that
the linear gradient typically used in liquid chromatography are sub-optimal, and that one should be able to replace such gradients with nonlinear functions that produce even
distributions of the peptides across runs.

We tested this hypothesis by calculating and evaluating two types of
nonlinear gradients \cite{gradopt1}. Our results showed that the new
gradients not only provided more uniform distributions of the
peptides, but also led to an increase in the number of peptide
identifications. In addition, they facilitated the identification of
different peptide species compared to the typical linear gradients. To
allow the easy calculation of nonlinear gradients by any MS
practitioner, we also implemented a user-friendly graphical user
interface that can be downloaded free of cost \cite{gradopt2}.  Seen
from a general perspective, this showed that the gradient function is
yet another parameter that researchers can alter in their experiments,
opening up a whole new range of potential applications.

>>>>

\begin{wrapfigure}{r}{.45\textwidth}
\vspace{-10pt}
%\begin{figure}[htbp]
\centering
\includegraphics[trim=0.5cm 0cm 2cm 1.5cm, clip=true, width=0.43\textwidth]{img/reproducibility.pdf}
\caption{\label{fig:repr} {\bf Reproducibility of peptide retention
    time.} Two identical shotgun proteomics experiments were carried
  out. The retention times of the peptides confidently identified in
  both runs are displayed.}
%\end{figure}
\vspace{-60pt}
\end{wrapfigure}

% iRT? \cite{Escher:2012qv}
Following these observations, a variety of retention time prediction
methods has been proposed. Most naturally one could divide these
methods into four classes: look-up tables-based approaches,
index-based methods, modeling-based methods and machine learning
techniques. The main ideas underlying each category are illustrated in
Figure \ref{fig:methods} , while Table \ref{tab:rtmethods} lists a few
of the most popular retention time prediction methods from each
category.

\begin{figure}[!h]
\centering \includegraphics[clip=true,
  width=0.8\textwidth]{img/methods.pdf}
\caption{\label{fig:methods} {\bf Retention time prediction methods.}
  Each figure illustrates a class of retention time prediction
  methods. The details of the different methods are described in
  detail in Section \ref{sec:rtpred}}.
\vspace{-7pt}
\end{figure}

@@

Moruz {\em et al.} \cite{elude1} developed a software package for
retention time prediction named {\sc Elude}. One of the main novelty
points of {\sc Elude} was the calculation of peptide features based on
both a general hydrophobicity index, as well as a retention index
trained for each dataset at hand. This ensured that the algorithm
could be applied to different chromatography setups with minimal loss
in performance. For cases when little training data was available,
{\sc Elude} implemented an additional workflow based on a library of
pre-trained retention models. This workflow required only a few tens
of training peptides to select and calibrate the most suitable model
from the library. Both the selection and calibration were performed
using statistical methods robust to outliers.

However, despite all these efforts, it should be pointed out that none
of the retention time prediction methods currently available is able
to provide accuracies comparable to the small shifts observed when
running a sample twice through a chromatography column
(Figure \ref{fig:repr}). Recently, the lack of suitable
representations for secondary structure elements was indicated as one
of the main causes of prediction errors \cite{Reimer2012}.  Most
probably additional factors are involved, and further endeavors are
required to reach the desired level of accuracy.

@@

To address this, in Moruz {\em et al.}\cite{elude2} the version of
their algorithm {\sc ELUDE} was extended to encompass modified
peptides as well. This was straightforward to achieve once the authors
observed that the in-house trained retention index contained nearly
all the information incorporated in the more general hydrophobicity
scale.  This meant that they could remove all the features based on
this index with minimal loss in accuracy. Further, they redefined all
features describing a peptide in terms of the in-house retention index
alone.  Since this index was trained for each dataset at hand, the
algorithm became fully trainable for any modified amino acid (Figure
\ref{fig:elude}). The new version of {\sc ELUDE} was evaluated on
peptides containing artifactual modifications introduced as part of an
MS protocol called COmbined FRActional DIagonal Chromatography
(COFRADIC) \cite{Gevaert2002}. The results showed that {\sc ELUDE}
yielded equally accurate predictions for modified and unmodified
peptides \cite{elude2}.

@@

% how it works and main results
More specifically, they matched observed masses and retention times of
unfragmented ions to predicted masses and retention times for the
theoretical peptides from an {\it in silico} digest. They built a
statistical framework to assign protein probabilities based on these
matches, and combined the results with the ones obtained by the
typical workflow based on fragmented ions. Following this approach, they
obtained up to 7\% more protein identifications at a fixed false
discovery rate threshold of 1\% \cite{mf}.

The idea of identifying proteins based on entire peptides, rather than
peptide fragments, is not new. In fact, this was the principle
underlying peptide mass fingerprinting, one of the first methods used
to identify proteins via mass spectrometry. However, when introduced
in the 90's, mass fingerprinting could not be applied to unequivocally
identify proteins in complex biological samples. Consequently, shotgun
proteomics has gained popularity instead. Nevertheless, with great
developments in instrumentation leading to sub-ppm mass accuracy, the
possibility of using mass fingerprinting-like methods came back to
attention~\cite{iamt1}. The mentioned study showed how improved mass
accuracy and retention time predictions enable the use of unfragmented
ions even in the context of complex peptide mixtures.

@@

% Improve the design of SRM experiments 
In addition, predicted retention times are useful investigating the
elution properties of the peptides.  To illustrate this, in
\cite{elude1} the retention time of all the theoretical peptides from
an {\it in silico} digest of the human proteome were predicted using
{\sc ELUDE}. This exercise revealed a rather skewed distribution, with
a majority of the peptides eluting in only a small portion of the
chromatography run. This suggested that the linear gradient typically
used in liquid chromatography are sub-optimal, and that one should be
able to replace such gradients with nonlinear functions that produce
even distributions of the peptides across runs.

This hypothesis was tested by calculating and evaluating two types of
nonlinear gradients \cite{gradopt1}. The results showed that the new
gradients not only provided more uniform distributions of the
peptides, but also led to an increase in the number of peptide
identifications. In addition, they facilitated the identification of
different peptide species compared to the typical linear gradients. To
allow the easy calculation of nonlinear gradients by any MS
practitioner, a user-friendly graphical user interface was implemented 
\cite{gradopt2}.  Seen from a general perspective, this showed that
the gradient function is yet another parameter that researchers can
alter in their experiments, opening up a whole new range of potential
applications.
